---
title: Detecting invasive mammals for management in Patagonia
subtitle: Obtaining nature media with the R package suwo
author: <font size="4"><a href="https://marce10.github.io/">Marcelo Araya-Salas, PhD</a>
date: "`r Sys.Date()`"
toc: true
toc-depth: 3
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: true
    code-tools: true
    css: qmd.css
editor_options:
  chunk_output_type: console
---

```{r set root directory}
#| eval: true
#| echo: false

# set working directory
knitr::opts_knit$set(root.dir =  "..")

```

```{r add link to github repo}
#| eval: true
#| output: asis
#| echo: false

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  if (length(url) > 0 && nchar(url) > 1)
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```

```{r setup style}
#| message: false
#| warning: false

# options to customize chunk outputs
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 65),
  tidy = TRUE,
  message = FALSE
 )

```

<!-- skyblue box -->

::: {.alert .alert-warning}

This analysis serves as the invasive-species-management case study accompanying the manuscript **"Obtaining nature media with the R package suwo"**, illustrating how suwo-retrieved images can be carried into a model-training pipeline.

:::

::: {.alert .alert-info}
# Purpose {.unnumbered .unlisted}

-   Assemble images of the focal management species -- American Mink (*Neogale vison*), Black Rat (*Rattus rattus*), and *Lontra* otters (*L. provocax*, *L. annectens*, *L. longicaudis*, *L. canadensis*, *L. felina*) -- from iNaturalist using the R package `suwo`.

-   Test whether pretrained YOLO models (trained on COCO classes) already detect the focal individuals well enough to bootstrap training labels.

-   Auto-generate YOLO-format training labels from those pretrained-model detections, remapped to the focal classes.

-   Train and evaluate a YOLO model to detect and classify the focal species, for use in an automated trap-release system targeting the invasive American Mink in Patagonia.

:::

- The model-training steps below are Python (`ultralytics`, `opencv-python`), run through `reticulate`. They assume a Python environment with those packages already installed (e.g. `pip install ultralytics opencv-python`) and, for training, a CUDA-capable GPU.

- Raw images are **not** included in this repository (tens of thousands of files, downloaded on demand). The query metadata returned by `suwo` **is** included (one CSV per species in `data/processed/mink_detection_queries/`, 74,490 records), so the data-availability plots and map below are rendered directly from it. Auto-generated labels and the train/validation split are not included -- they are reproducible from the chunks below, which default to `eval: false` since every step there is either a slow bulk download or requires a GPU. The one artifact kept for reference is the log of an actual download run: [`data/raw/mink_detection_logs/download_images.log`](../data/raw/mink_detection_logs/download_images.log) ([download](https://github.com/maRce10/suwo_publication/raw/main/data/raw/mink_detection_logs/download_images.log); 1,131 of 1,461 files downloaded successfully in that run, ~3.9 hours).

# Analysis flowchart {.unnumbered .unlisted}

```{mermaid}

flowchart LR
  A[Query & download images] --> B("Preview</br>pretrained-model</br>detections")
  B --> C(Auto-generate labels)
  C --> D(Train / validation split)
  D --> E(Train YOLO model)
  E --> F(Evaluate)

style A fill:#382A5433
style B fill:#395D9C33
style C fill:#3497A933
style D fill:#60CEAC33
style E fill:#60CEAC33
style F fill:#60CEAC33

```

# Load packages {.unnumbered .unlisted}

```{r load packages}
#| eval: true
#| warning: false

# if sketchy not installed install it
if (!require("sketchy")) {
  install.packages("sketchy")
}

# install/ load packages

sketchy::load_packages(
  packages = c(
    "knitr",
    "tidyverse",
    "ggpattern",
    "maps",
    github = "maRce10/suwo",
    "reticulate"
  )
)

```

# Query and download images

```{r query and download images}
#| eval: false

# Folder layout, relative to the project root
IMAGES_FOLDER  <- "./data/raw/mink_detection_images"
QUERIES_FOLDER <- "./data/processed/mink_detection_queries"

dir.create(IMAGES_FOLDER, showWarnings = FALSE, recursive = TRUE)
dir.create(QUERIES_FOLDER, showWarnings = FALSE, recursive = TRUE)

#~~~~~~~~~~~~~~~~
# Queries
#~~~~~~~~~~~~~~~~
# American Mink
neogale_vison <- query_inaturalist(species = "Neogale vison", format = "image")  # ~12 minutes
write.csv(neogale_vison, paste0(QUERIES_FOLDER, "/neogale_vison.csv"), row.names = FALSE)

# Rat
rattus_rattus <- query_inaturalist(species = "Rattus rattus", format = "image")  # ~25 minutes
write.csv(rattus_rattus, paste0(QUERIES_FOLDER, "/rattus_rattus.csv"), row.names = FALSE)

# Lontras: L. provocax, L. annectens, L. longicaudis, L. canadensis, L. felina
lontra_provocax <- query_inaturalist(species = "Lontra provocax", format = "image")
write.csv(lontra_provocax, paste0(QUERIES_FOLDER, "/lontra_provocax.csv"), row.names = FALSE)

lontra_annectens <- query_inaturalist(species = "Lontra annectens", format = "image")
write.csv(lontra_annectens, paste0(QUERIES_FOLDER, "/lontra_annectens.csv"), row.names = FALSE)

lontra_longicaudis <- query_inaturalist(species = "Lontra longicaudis", format = "image")
write.csv(lontra_longicaudis, paste0(QUERIES_FOLDER, "/lontra_longicaudis.csv"), row.names = FALSE)

lontra_canadensis <- query_inaturalist(species = "Lontra canadensis", format = "image")  # ~17 minutes
write.csv(lontra_canadensis, paste0(QUERIES_FOLDER, "/lontra_canadensis.csv"), row.names = FALSE)

lontra_felina <- query_inaturalist(species = "Lontra felina", format = "image")
write.csv(lontra_felina, paste0(QUERIES_FOLDER, "/lontra_felina.csv"), row.names = FALSE)

#~~~~~~~~~~~~~~~~
# Download images
#~~~~~~~~~~~~~~~~
N_CORES <- 20

neogale_vison <- read.csv(paste0(QUERIES_FOLDER, "/neogale_vison.csv"))
download_media(metadata = neogale_vison, cores = N_CORES, path = IMAGES_FOLDER)

rattus_rattus <- read.csv(paste0(QUERIES_FOLDER, "/rattus_rattus.csv"))
download_media(metadata = rattus_rattus, cores = N_CORES, path = IMAGES_FOLDER)

lontra_provocax <- read.csv(paste0(QUERIES_FOLDER, "/lontra_provocax.csv"))
download_media(metadata = lontra_provocax, cores = N_CORES, path = IMAGES_FOLDER)

lontra_annectens <- read.csv(paste0(QUERIES_FOLDER, "/lontra_annectens.csv"))
download_media(metadata = lontra_annectens, cores = N_CORES, path = IMAGES_FOLDER)

lontra_longicaudis <- read.csv(paste0(QUERIES_FOLDER, "/lontra_longicaudis.csv"))
download_media(metadata = lontra_longicaudis, cores = N_CORES, path = IMAGES_FOLDER)

lontra_canadensis <- read.csv(paste0(QUERIES_FOLDER, "/lontra_canadensis.csv"))
download_media(metadata = lontra_canadensis, cores = N_CORES, path = IMAGES_FOLDER)

lontra_felina <- read.csv(paste0(QUERIES_FOLDER, "/lontra_felina.csv"))
download_media(metadata = lontra_felina, cores = N_CORES, path = IMAGES_FOLDER)


```

# Visualize data availability

* Query metadata (saved above, one CSV per species) is combined to show how many images are available per species, and the same counts collapsed to `Lontra spp.` given how few images each individual otter species has relative to the mink and rat.

```{r data availability barplots}
#| eval: true
#| message: false

queries_files <- list.files(
  path = "./data/processed/mink_detection_queries",
  pattern = "*.csv",
  full.names = TRUE
)

combined_data <- bind_rows(lapply(queries_files, read_csv, show_col_types = FALSE))
combined_data %>% glimpse()

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# barplot of number of images per species
combined_data %>%
  group_by(species) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = species, y = n)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("./output/mink_detection_species_barplot.png",
       width = 1200, height = 800,
       scale = 1.2, units = "px")

# colors for plots
sp_colors <- adjustcolor(c("red", "gray", "blue", "green", "orange", "purple", "darkred"),alpha.f = 0.7)

sp_colors <- viridis::mako(7, alpha = 0.9)
names(sp_colors) <- c("Neogale vison", "Rattus rattus", "Lontra provocax", "Lontra longicaudis",
                       "Lontra felina", "Lontra canadensis", "Lontra annectens")

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# same, but stack species with "Lontra" in the name as "Lontra spp."
combined_data %>%
  mutate(species1 = ifelse(grepl("Lontra", species), "Lontra spp.", species)) %>%
  mutate(species = factor(species, levels = c(
    "Neogale vison", "Rattus rattus", "Lontra provocax", "Lontra longicaudis",
    "Lontra felina", "Lontra canadensis", "Lontra annectens"
  ))) %>%
  group_by(species1, Species = species) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = species1, y = n, fill = Species)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = sp_colors,
    breaks = c("Lontra provocax", "Lontra longicaudis", "Lontra felina",
               "Lontra canadensis", "Lontra annectens")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "")

ggsave("./output/mink_detection_species_barplot_lontra_spp.png",
       width = 1200, height = 800,
       scale = 1.2, units = "px")

```

* Number of detections actually used for training (produced by the YOLO-label step below), per class:

```{r training data barplot}
#| eval: false

# training data: detections actually used after the YOLO-label step below
counts <- read.csv("./data/processed/mink_detection_class_detection_counts.csv")

counts %>%
  mutate(Species = case_when(
    class_name == "Neogale_vison" ~ "Neogale vison",
    class_name == "Rattus_rattus" ~ "Rattus rattus",
    class_name == "Lontra" ~ "Lontra spp."
  )) %>%
  rename(n = detection_count) %>%
  ggplot(aes(x = Species, y = n)) +
  geom_bar(stat = "identity") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("./output/mink_detection_training_data_barplot.png",
       width = 1200, height = 800,
       scale = 1.2, units = "px")

```

# Map record locations

* Geographic distribution of the retrieved records, colored by species. `suwo::map_locations()` builds an interactive map (with clustering and photo popups); with ~74k records the resulting widget is too heavy (~100 MB) to embed in this page, so the code is shown but not run here, and a static version of the same map is rendered below instead.

```{r interactive map of records}
#| eval: false

combined_data$species <- factor(
  combined_data$species,
  levels = c("Neogale vison", "Rattus rattus", "Lontra provocax", "Lontra longicaudis",
             "Lontra felina", "Lontra canadensis", "Lontra annectens"),
  ordered = TRUE
)

map_locations(
  combined_data,
  cluster = TRUE,
  by = "species",
  type = "circles",
  palette = function(n) rep(sp_colors,
                            length.out = n),
  tags = c("repository", "key", "species", "date", "country", "locality", "user_name"),
  show_media = TRUE,
  popup_size = 1
)
```

```{r static map of records}
#| eval: true
#| fig-width: 10
#| fig-height: 5.5

species_levels <- c("Neogale vison", "Rattus rattus", "Lontra provocax", "Lontra longicaudis",
                    "Lontra felina", "Lontra canadensis", "Lontra annectens")

map_data_df <- combined_data %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  mutate(species = factor(species, levels = species_levels))

ggplot() +
  geom_polygon(
    data = map_data("world"),
    aes(x = long, y = lat, group = group),
    fill = "gray92", color = "gray70", linewidth = 0.1
  ) +
  geom_point(
    data = map_data_df,
    aes(x = longitude, y = latitude, color = species),
    size = 0.4, alpha = 0.5
  ) +
  scale_color_manual(
    values = sp_colors,
    name = "Species"
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  coord_quickmap(ylim = c(-60, 85), expand = FALSE) +
  theme_light() +
  theme(legend.text = element_text(face = "italic")) +
  labs(x = "Longitude", y = "Latitude")

ggsave("./output/mink_detection_records_map.png",
       width = 10, height = 5.5, dpi = 300)
```

# Preview pretrained-model detections

* Before training a dedicated model, a pretrained YOLO model (trained on COCO classes) is run on a sample of the downloaded images to check whether any COCO class already picks up the focal individuals (the manuscript reports this for the `bird`, `dog`, `cat`, and `bear` classes across YOLO v8, v10, and 26).

```{python preview yolo detections}
#| eval: false

import cv2
from pathlib import Path
from ultralytics import YOLO
import time

# Paths (relative to the project root)
IMAGE_DIR = Path("./data/raw/mink_detection_images")
OUTPUT_DIR = Path("./output/mink_detection_test_inference")

# Pretrained YOLO model
MODEL_NAME = "yolov8l.pt"

OUTPUT_DIR.mkdir(parents=True, exist_ok=True)

def main():
    start = time.perf_counter()

    print(f"Loading pretrained YOLO model: {MODEL_NAME}")
    model = YOLO(MODEL_NAME)

    image_extensions = {".jpg", ".jpeg", ".png", ".bmp", ".gif", ".webp"}
    image_files = [f for f in IMAGE_DIR.iterdir()
                   if f.is_file() and f.suffix.lower() in image_extensions]

    print(f"Found {len(image_files)} images to process")

    for idx, image_path in enumerate(image_files, 1):
        print(f"[{idx}/{len(image_files)}] Processing: {image_path.name}")
        try:
            results = model(str(image_path), conf=0.5)
            annotated_frame = results[0].plot()
            output_path = OUTPUT_DIR / image_path.name
            cv2.imwrite(str(output_path), annotated_frame)
        except Exception as e:
            print(f"Error processing {image_path.name}: {e}")
            continue

    print(f"\nInference complete! Labeled images saved to: {OUTPUT_DIR}")
    print(f"Total images processed: {len(image_files)}")

    elapsed = time.perf_counter() - start
    print(f"Runtime: {elapsed:.2f} seconds")

if __name__ == "__main__":
    main()

```

# Auto-generate YOLO training labels

* Detections from the pretrained model are kept only for the COCO classes that approximate the focal species (`bird`, `dog`, `cat`, `bear`) and remapped to the three focal classes used for training: `0` = American Mink (`Neogale_vison`), `1` = Black Rat (`Rattus_rattus`), `2` = *Lontra* spp. (all `Lontra_*` species pooled). A label file is written only when at least one valid detection survives this filter, which is why a large fraction of the downloaded images end up unused for training (see manuscript for exact counts).

```{python auto-generate yolo labels}
#| eval: false

from pathlib import Path
from ultralytics import YOLO
import argparse
import time

# Paths (relative to the project root)
IMAGE_DIR = Path("./data/raw/mink_detection_images")
LABEL_DIR = Path("./data/processed/mink_detection_labels")

# Pretrained model
MODEL_NAME = "yolov8l.pt"

# Allowed detector classes from COCO
ALLOWED_DETECTION_CLASSES = {"bird", "dog", "cat", "bear"}

# Species-to-output-class mapping from filename prefix
SPECIES_TO_TARGET_CLASS = {
    "Neogale_vison": 0,
    "Rattus_rattus": 1,
    "Lontra_provocax": 2,
    "Lontra_annectens": 2,
    "Lontra_longicaudis": 2,
    "Lontra_canadensis": 2,
    "Lontra_felina": 2,
}

IMAGE_EXTENSIONS = {".jpg", ".jpeg", ".png", ".bmp", ".gif", ".webp"}


def parse_args():
    parser = argparse.ArgumentParser(description="Run YOLO inference and export filtered YOLO labels.")
    parser.add_argument(
        "--limit",
        type=int,
        default=None,
        help="Optional integer limit for number of images to process. If omitted, all images are used."
    )
    parser.add_argument(
        "--conf",
        type=float,
        default=0.5,
        help="Confidence threshold for YOLO inference."
    )
    return parser.parse_args()


def get_species_from_filename(image_path):
    """Expected filename format: species-id.extension, e.g. Rattus_rattus-INAT383396732.jpeg"""
    species = image_path.stem.split("-", 1)[0]
    return species if species in SPECIES_TO_TARGET_CLASS else None


def main():
    args = parse_args()
    start = time.perf_counter()

    LABEL_DIR.mkdir(parents=True, exist_ok=True)

    print(f"Loading pretrained YOLO model: {MODEL_NAME}")
    model = YOLO(MODEL_NAME)

    model_name_to_id = {name: idx for idx, name in model.names.items()}
    valid_model_class_ids = {
        model_name_to_id[name]
        for name in ALLOWED_DETECTION_CLASSES
        if name in model_name_to_id
    }

    print(f"Detector classes used: {sorted(ALLOWED_DETECTION_CLASSES)}")
    print(f"Resolved detector class ids: {sorted(valid_model_class_ids)}")

    image_files = sorted(
        f for f in IMAGE_DIR.iterdir()
        if f.is_file() and f.suffix.lower() in IMAGE_EXTENSIONS
    )

    if args.limit is not None:
        if args.limit < 1:
            raise ValueError("--limit must be a positive integer")
        image_files = image_files[:args.limit]

    print(f"Found {len(image_files)} images to process")

    processed = 0
    exported = 0
    skipped_unknown_species = 0
    no_valid_detection = 0

    for idx, image_path in enumerate(image_files, 1):
        print(f"[{idx}/{len(image_files)}] Processing: {image_path.name}")

        species = get_species_from_filename(image_path)
        if species is None:
            print("  Skipped: unknown species prefix in filename")
            skipped_unknown_species += 1
            continue

        target_class = SPECIES_TO_TARGET_CLASS[species]

        try:
            results = model.predict(source=str(image_path), conf=args.conf, verbose=False)
            result = results[0]
            boxes = result.boxes

            output_lines = []

            if boxes is not None and len(boxes) > 0:
                cls_list = boxes.cls.tolist()
                xywhn_list = boxes.xywhn.tolist()

                for cls_id, xywhn in zip(cls_list, xywhn_list):
                    cls_id = int(cls_id)
                    if cls_id not in valid_model_class_ids:
                        continue
                    x_center, y_center, width, height = xywhn
                    output_lines.append(
                        f"{target_class} {x_center:.6f} {y_center:.6f} {width:.6f} {height:.6f}"
                    )

            if output_lines:
                label_path = LABEL_DIR / f"{image_path.stem}.txt"
                with open(label_path, "w", encoding="utf-8") as f:
                    f.write("\n".join(output_lines) + "\n")
                exported += 1
                print(f"  Exported: {label_path.name} ({len(output_lines)} labels)")
            else:
                no_valid_detection += 1
                print("  No valid detections (bird/dog/cat/bear), no label exported")

            processed += 1

        except Exception as e:
            print(f"  Error processing {image_path.name}: {e}")
            continue

    elapsed = time.perf_counter() - start

    print("\nInference complete")
    print(f"Images considered: {len(image_files)}")
    print(f"Images processed: {processed}")
    print(f"Label files exported: {exported}")
    print(f"Skipped due to unknown species prefix: {skipped_unknown_species}")
    print(f"Images with no valid detections: {no_valid_detection}")
    print(f"YOLO labels saved to: {LABEL_DIR}")
    print(f"Runtime: {elapsed:.2f} seconds")


if __name__ == "__main__":
    main()

```

# Train / validation split

* Images with an exported label are randomly split 80/20 into training and validation sets (fixed seed for reproducibility), and copied into the layout YOLO expects.

```{python train validation split}
#| eval: false

from pathlib import Path
import random
import shutil

src_labels = Path("./data/processed/mink_detection_labels")
src_images = Path("./data/raw/mink_detection_images")
dst_root = Path("./data/processed/mink_detection_yolo_split")

train_ratio = 0.8
seed = 42
image_exts = [".jpg", ".jpeg", ".png", ".bmp", ".tif", ".tiff", ".webp"]

def find_image(stem):
    for ext in image_exts:
        candidate = src_images / f"{stem}{ext}"
        if candidate.exists():
            return candidate
    matches = []
    for ext in image_exts:
        matches.extend(src_images.glob(f"**/{stem}{ext}"))
    return matches[0] if matches else None

def main():
    if not src_labels.exists():
        raise FileNotFoundError(f"Labels folder not found: {src_labels}")
    if not src_images.exists():
        raise FileNotFoundError(f"Images folder not found: {src_images}")

    label_files = sorted(src_labels.glob("*.txt"))
    pairs = []
    missing = []

    for label_path in label_files:
        image_path = find_image(label_path.stem)
        if image_path is None:
            missing.append(label_path.name)
            continue
        pairs.append((label_path, image_path))

    if not pairs:
        raise RuntimeError("No label-image pairs found.")

    random.seed(seed)
    random.shuffle(pairs)

    split_idx = int(len(pairs) * train_ratio)
    train_pairs = pairs[:split_idx]
    val_pairs = pairs[split_idx:]

    for split_name, split_pairs in [("train", train_pairs), ("val", val_pairs)]:
        (dst_root / "labels" / split_name).mkdir(parents=True, exist_ok=True)
        (dst_root / "images" / split_name).mkdir(parents=True, exist_ok=True)
        for label_path, image_path in split_pairs:
            shutil.copy2(label_path, dst_root / "labels" / split_name / label_path.name)
            shutil.copy2(image_path, dst_root / "images" / split_name / image_path.name)

    print(f"Total pairs: {len(pairs)}")
    print(f"Train: {len(train_pairs)}")
    print(f"Val: {len(val_pairs)}")

    if missing:
        print(f"Missing images for {len(missing)} labels:")
        for name in missing[:20]:
            print(f"  {name}")
        if len(missing) > 20:
            print("  ...")

if __name__ == "__main__":
    main()

```

# Dataset configuration

* YOLO's training command (next section) points at a small YAML file declaring the three focal classes and the train/validation image directories produced above: [`scripts/mink_detection_data.yaml`](mink_detection_data.yaml) ([download](https://github.com/maRce10/suwo_publication/raw/main/scripts/mink_detection_data.yaml)).

```{r show yaml}
#| echo: false
#| eval: true
#| comment: ""

cat(readLines("./scripts/mink_detection_data.yaml"), sep = "\n")

```

# Train the YOLO model

* The YOLO v10 nano model is trained from the split above. This step needs a CUDA-capable GPU and was not re-run as part of knitting this document; see the manuscript for the training hardware, runtime, and resulting accuracy metrics (confusion matrix, mAP50, precision).

```{bash train yolo model}
#| eval: false

yolo detect train data=./scripts/mink_detection_data.yaml model=yolov10l.pt epochs=100 imgsz=640 batch=16

```

---

# References {.unnumbered .unlisted}

- Redmon, J., Divvala, S., Girshick, R., & Farhadi, A. (2016). You only look once: Unified, real-time object detection. *CVPR*.

- Jocher, G., Qiu, J., & Chaurasia, A. (2023). Ultralytics YOLO (Version 8.0.0) [Computer software]. https://github.com/ultralytics/ultralytics

# Session information {.unnumbered .unlisted}

<details>
  <summary>Click to see</summary>
```{r session info}
#| echo: false

if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::session_info()
} else {
  sessionInfo()
}

```
</details>
